---
title: "SARS-CoV-2 6 month T cells"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(SingleCellExperiment)
library(scater)
library(DropletUtils)
library(scran)

```

## Analysis of 10x 6 month convalescent Dong050121 tube 2 samples using Single Cell Experiment
Following this guide http://bioconductor.org/books/release/OSCA/droplet-processing.html

### Calling cells from empty droplets
```{r, fig.width=6, fig.height=4}
# read in data and create Single Cell Experiment object
# have to use raw unfiltered matrix if going to do cell calling from scratch
# without using CellRanger's cell calling
t2 <- read10xCounts("/t1-data/user/lfelce/10x_Dong050121/combined_T2_repeat4/outs/raw_feature_bc_matrix", type="sparse")

bcrank <- barcodeRanks(counts(t2))

# Only showing unique points for plotting speed.
uniq <- !duplicated(bcrank$rank)
plot(bcrank$rank[uniq], bcrank$total[uniq], log="xy",
    xlab="Rank", ylab="Total UMI count", cex.lab=1.2)

abline(h=metadata(bcrank)$inflection, col="darkgreen", lty=2)
abline(h=metadata(bcrank)$knee, col="dodgerblue", lty=2)

legend("bottomleft", legend=c("Inflection", "Knee"), 
        col=c("darkgreen", "dodgerblue"), lty=2, cex=1.2)

```
Total UMI count for each barcode in dataset, plotted against its rank (in decreasing order of total counts). Inferred locations of inflection and knee points are also shown.

### Testing for empty droplets
Check if expression profile for each cell barcode is significantly different from ambient RNA pool - can discriminate between well-sequenced empty droplets and droplets derived from cells with litle RNA. Any significant deviation indicates that barcode corresponds to cell-containing droplet. Use FDR 0.1% - no more than 0.1% of called barcodes should be empty droplets on average. 
```{r, fig.width=6, fig.height=4}
# emptyDrops performs Monte Carlo simulations to compute p-values,
# so we need to set the seed to obtain reproducible results.
set.seed(100)
e.out <- emptyDrops(counts(t2))

# See ?emptyDrops for an explanation of why there are NA values.
summary(e.out$FDR <= 0.001)

table(Sig=e.out$FDR <= 0.001, Limited=e.out$Limited)

# emptyDrops() assumes barcodes with low total UMI counts are empty droplets
# look at distribution of p-values for low-total barcode - should have uniform distribution
# large peaks near 0 indicate barcodes with total counts below lower are not all ambient in origin
# can decrease lower further to ensure that barcodes corresponding to droplets with very small cell counts are not used to estimate ambient profile

set.seed(100)
limit <- 100   
all.out <- emptyDrops(counts(t2), lower=limit, test.ambient=TRUE)
hist(all.out$PValue[all.out$Total <= limit & all.out$Total > 0],
    xlab="P-value", main="", col="grey80") 

# subset SingleCellExperiment object to only retain detected cells
t2 <- t2[,which(e.out$FDR <= 0.001)]

```

### Filtering on mitochondrial proportion
Check that not removing subpopulation of metabolically active cells. May want to do alongside emptyDrops() to remove non-empty droplets containing cell fragments or stripped nuclei not caught by mitochdondrial filter.
```{r, fig.width=6, fig.height=4}
is.mito <- grep("^MT-", rowData(t2)$Symbol)
pbmc.qc <- perCellQCMetrics(t2, subsets=list(MT=is.mito))
discard.mito <- isOutlier(pbmc.qc$subsets_MT_percent, type="higher")
summary(discard.mito)

plot(pbmc.qc$sum, pbmc.qc$subsets_MT_percent, log="x",
    xlab="Total count", ylab='Mitochondrial %')
abline(h=attr(discard.mito, "thresholds")["higher"], col="red")
# threshold should be set at ~ 10% not 0%?
```

### Removing ambient contamination
For routine analyses, there is usually no need to remove the ambient contamination from each library. A consistent level of contamination across the dataset does not introduce much spurious heterogeneity, so dimensionality reduction and clustering on the original (log-)expression matrix remain valid. For genes that are highly abundant in the ambient solution, we can expect some loss of signal due to shrinkage of the log-fold changes between clusters towards zero, but this effect should be negligible for any genes that are so strongly upregulated that they are able to contribute to the ambient solution in the first place. This suggests that ambient removal can generally be omitted from most analyses, though we will describe it here regardless as it can be useful in specific situations.

Effective removal of ambient contamination involves tackling a number of issues. We need to know how much contamination is present in each cell, which usually requires some prior biological knowledge about genes that should not be expressed in the dataset (e.g., mitochondrial genes in single-nuclei datasets, see Section 19.4) or genes with mutually exclusive expression profiles (Young and Behjati 2018). Those same genes must be highly abundant in the ambient solution to have enough counts in each cell for precise estimation of the scale of the contamination. The actual subtraction of the ambient contribution also must be done in a manner that respects the mean-variance relationship of the count data. Unfortunately, these issues are difficult to address for single-cell data due to the imprecision of low counts,

Rather than attempting to remove contamination from individual cells, a more measured approach is to operate on clusters of related cells. The removeAmbience() function from DropletUtils will remove the contamination from the cluster-level profiles and propagate the effect of those changes back to the individual cells. Specifically, given a count matrix for a single sample and its associated ambient profile, removeAmbience() will:

1. Aggregate counts in each cluster to obtain an average profile per cluster.

2. Estimate the contamination proportion in each cluster with maximumAmbience() (see Section 14.4). This has the useful property of not requiring any prior knowledge of control or mutually exclusive expression profiles, albeit at the cost of some statistical rigor.

3. Subtract the estimated contamination from the cluster-level average.

4. Perform quantile-quantile mapping of each individual cell’s counts from the old average to the new subtracted average. This preserves the mean-variance relationship while yielding corrected single-cell profiles.
```{r}
# this randomly repeats all of the above ...

# cell detection
set.seed(100)
e.out <- emptyDrops(counts(t2))
t2 <- t2[,which(e.out$FDR <= 0.001)]

# quality control
stats <- perCellQCMetrics(t2, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
t2 <- t2[,!high.mito]

# normalisation
set.seed(1000)
clusters <- quickCluster(t2)
t2 <- computeSumFactors(t2, cluster=clusters)
t2 <- logNormCounts(t2)

# variance modelling
set.seed(1001)
dec.t2 <- modelGeneVarByPoisson(t2)
top.t2 <- getTopHVGs(dec.pbmc, prop=0.1)

# dimensionality reduction
set.seed(10000)
t2 <- denoisePCA(t2, subset.row=top.t2, technical=dec.t2)

set.seed(100000)
t2 <- runTSNE(t2, dimred="PCA")

set.seed(1000000)
t2 <- runUMAP(t2, dimred="PCA")

# clustering
g <- buildSNNGraph(t2, k=10, use.dimred = 'PCA')
clust <- igraph::cluster_walktrap(g)$membership
colLabels(t2) <- factor(clust)

# Not all genes are reported in the ambient profile from emptyDrops,
# as genes with counts of zero across all droplets are just removed.
# So for convenience, we will restrict our analysis to genes with 
# non-zero counts in at least one droplet (empty or otherwise).
amb <- metadata(e.out)$ambient[,1]
stripped <- t2[names(amb),]

out <- removeAmbience(counts(stripped), ambient=amb, groups=colLabels(stripped))
dim(out)
```

