---
title: "SARS-CoV-2 6 month T cells"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Seurat)
library(plyr)
library(dplyr)
library(patchwork)
library(cowplot)
library(ggplot2)
library(data.table)
library(tidyverse)
library(harmony)
library(gridExtra)
```

## 10x single cell RNA sequencing from SARS-CoV-2 CD4/8 tetramer sorted T cells

### Batch 1
Dong 171220 = batch1
Hashtag Ab 1 1180_6Mon, expanded, DR0101_M24 tetramer, CD4
Hashtag Ab 2 1344_6Mon, expanded, DR1501_S51 tetramer, CD4
Hashtag Ab 4 1344_6Mon, expanded, DR1501_S174 tetramer, CD4
Hashtag Ab 5 1180_6Mon, expanded, A3_S76 tetramer, CD8
Hashtag Ab 6 1180_6Mon, expanded, A1-ORF3a-28 tetramer, CD8
Hashtag Ab 7 1344_6Mon, expanded, B27_NP1 tetramer, CD8
Hashtag Ab 8 1344_6Mon, expanded, B27_NP13 tetramer, CD8
Note: hashtag 5 and 6 mixed up?

```{r}
# Read in data
batch1.counts <- Read10X(data.dir = "/t1-data/user/lfelce/10x_DONG171220/counts_gex/filtered_feature_bc_matrix")
colnames(batch1.counts) <- gsub(x = colnames(batch1.counts), pattern = "-1", replacement = "")

batch1.hash <- Read10X(data.dir ="/t1-data/user/lfelce/10x_DONG171220/counts_ab/filtered_feature_bc_matrix")
colnames(batch1.hash) <- gsub(x = colnames(batch1.hash), pattern = "-1", replacement = "")

# Select cell barcodes detected by both RNA and hashtag, then filter
joint.barcodes <- intersect(colnames(batch1.counts), colnames(batch1.hash))

# Subset RNA and hashtag counts by joint cell barcodes
batch1.counts <- batch1.counts[, joint.barcodes]
batch1.hash <- as.matrix(batch1.hash[,joint.barcodes])

# remove Hashtag3 (no cells labelled with this Ab)
batch1.hash <- batch1.hash[-3,]

# rename rows to sample names
rownames(batch1.hash) <- c("1180_CD4_M24", "1344_CD4_S51", "1344_CD4_S174", "1180_CD8_S76", "1180_CD8_ORF3a-28", "1344_CD8_NP1", "1344_CD8_NP13")

# counts per hashtag
rowSums(batch1.hash)
rowMeans(batch1.hash)

# Set up Seurat object 
batch1 <- CreateSeuratObject(counts = batch1.counts)

# Normalize RNA data with log normalization
batch1 <- NormalizeData(batch1)

# Find and scale variable features
batch1 <- FindVariableFeatures(batch1, selection.method = "mean.var.plot")

batch1 <- ScaleData(batch1, features = VariableFeatures(batch1))

# Add HTO data as a new assay independent from RNA
batch1[["hashtag"]] <- CreateAssayObject(counts = batch1.hash)

# Normalize HTO data, here we use centered log-ratio (CLR) transformation
batch1 <- NormalizeData(batch1, assay = "hashtag", normalization.method = "CLR")

# Demultiplex cells based on hashtag enrichment - use HTODemux() to assign single cells back to sample origins

# If you have a very large dataset we suggest using k_function = 'clara'. This is a k-medoid
# clustering function for large applications You can also play with additional parameters (see
# documentation for HTODemux()) to adjust the threshold for classification Here we are using the
# default settings
batch1 <- HTODemux(batch1, assay = "hashtag", positive.quantile = 0.99)

```

#### Visualise demultiplexing results
Output from running HTODemux is saved in metadata. Can visualise cells classified as singletse, doublets and negative/ambiguous cells

```{r, fig.width=10, fig.height=12}
# Global classification results
table(batch1$hashtag_classification.global)

# Visualise enrichment for selected hashtags with ridge plots
# Group cells based on max hashtag signal
Idents(batch1) <- "hashtag_maxID"
RidgePlot(batch1, assay="hashtag", features=rownames(batch1[["hashtag"]]), ncol=2)

```

```{r, fig.width=6, fig.height=4}
# Compare number of UMIs for singlets, doublets and negative cells
Idents(batch1) <- "hashtag_classification.global"
VlnPlot(batch1, features = "nCount_RNA", pt.size = 0.1, log = TRUE)
```

```{r, fig.width=7, fig.height=4}
# Cluster and visualize cells using the usual scRNA-seq workflow, and examine for the potential presence of batch effects.

# Extract the singlets
batch1.singlet <- subset(batch1, idents = "Singlet")

# Select the top 1000 most variable features
batch1.singlet <- FindVariableFeatures(batch1.singlet, selection.method = "mean.var.plot")

# Scaling RNA data
batch1.singlet <- ScaleData(batch1.singlet, features = VariableFeatures(batch1.singlet))

# Run PCA
batch1.singlet <- RunPCA(batch1.singlet, features = VariableFeatures(batch1.singlet))

# Elbow plot
batch1.singlet <- JackStraw(batch1.singlet, num.replicate = 100)
batch1.singlet <- ScoreJackStraw(batch1.singlet, dims = 1:20)
JackStrawPlot(batch1.singlet, dims = 1:20)
```

```{r, fig.width=18, fig.height=10}
# We select the top 10 PCs for clustering 
batch1.singlet <- FindNeighbors(batch1.singlet, reduction = "pca", dims = 1:20)
batch1.singlet <- FindClusters(batch1.singlet, resolution = 0.5, verbose = FALSE)
batch1.singlet <- RunUMAP(batch1.singlet, dims = 1:20)
batch1.singlet <- RunTSNE(batch1.singlet, reduction = "pca", dims = 1:20)

# Projecting singlet identities on UMAP visualization
umap <- DimPlot(batch1.singlet, reduction="umap", group.by = "hashtag_classification")
tsne <- DimPlot(batch1.singlet, reduction="tsne", group.by = "hashtag_classification")
pca <- DimPlot(batch1.singlet, reduction="pca", group.by="hashtag_classification")
grid.arrange(umap, tsne, pca, nrow=2)


# Assign metadata to object
patient_list <- as.data.frame(batch1.singlet@meta.data$hashtag_classification)
colnames(patient_list) <- "sample"
patient_list$patient <- ifelse(grepl("1180", patient_list$sample), "1180",
                               ifelse(grepl("1344", patient_list$sample), "1344", ""))
patient_list$subset <- ifelse(grepl("CD4", patient_list$sample), "CD4",
                               ifelse(grepl("CD8", patient_list$sample), "CD8", ""))
patient_list$epitope <- ifelse(grepl("M24", patient_list$sample), "membrane",
                        ifelse(grepl("S51", patient_list$sample), "spike", 
                        ifelse(grepl("S174", patient_list$sample), "spike", 
                        ifelse(grepl("S76", patient_list$sample), "spike", 
                        ifelse(grepl("ORF3a-28", patient_list$sample), "ORF3a",
                        ifelse(grepl("NP1", patient_list$sample), "nucleoprotein", 
                        ifelse(grepl("NP13", patient_list$sample), "nucleoprotein", "")))))))

batch1.singlet@meta.data$"epitope" <- as.factor(patient_list$epitope)
batch1.singlet@meta.data$"patient" <- as.factor(patient_list$patient)
batch1.singlet@meta.data$"t-cell" <- as.factor(patient_list$subset)


# Save batch1 object
saveRDS(batch1.singlet, "batch1_seurat.rds")
```
#### Harmony integration
```{r, fig.height=3, fig.width, 6}
# Load in RDS
batch1 <- readRDS("batch1_seurat.rds")

batch1 <- batch1 %>%
    RunHarmony("t-cell", plot_convergence = TRUE)
# Harmony converged after 2 iterations

# access Harmony embeddings - show top 5 rows and top 5 columns
harmony_embeddings <- Embeddings(batch1, 'harmony')

```

```{r, fig.height=5, fig.width=14}
p1 <- DimPlot(object = batch1, reduction = "harmony", group.by = "hashtag_classification")
p2 <- VlnPlot(object = batch1, features = "harmony_1", group.by = "hashtag_classification", pt.size = .1)
p1 + p2
```

```{r, fig.height=5, fig.width=16}
p3 <- DimPlot(object = batch1, reduction = "pca", group.by="hashtag_classification")
p3 + pca
```
```{r, fig.width=16, fig.height=10}
# # use split.by="seurat_clusters" to also show UMAP by epitope and cluster number
# UMAP_epitope

batch1 <- batch1 %>%
    RunUMAP(reduction = "harmony", dims = 1:30) %>%
    FindNeighbors(reduction = "harmony", dims = 1:30) %>%
    FindClusters(resolution = 0.5) %>%
    identity()
umap_patient <- DimPlot(batch1, reduction = "umap", group.by = "patient")
umap_subset <- DimPlot(batch1, reduction = "umap", group.by = "t-cell")
umap_cluster <- DimPlot(batch1, reduction = "umap", group.by = "seurat_clusters")

grid.arrange(umap_patient, umap_subset, umap_cluster, nrow=2)
```


#### Combined GEX and Antibody Hashtag
Combined by putting two rows in libraries.csv for Feature Barcode analysis (cellranger count) for library_type Gene Expression and Antibody Capture. Followed 
https://satijalab.org/seurat/articles/multimodal_vignette.html to create Seurat object with two assays - RNA and HTO, but doesn't explain how to demultiplex, so only suitable for CITE-Seq.
```{r}
# # Read in data
# batch1.counts <- Read10X(data.dir = "/t1-data/user/lfelce/10x_DONG171220/counts_combined/filtered_feature_bc_matrix")
# 
# # Remove -1 from barcodes
# colnames(batch1.counts$`Gene Expression`) <- gsub(x = colnames(batch1.counts$`Gene Expression`), pattern = "-1", replacement = "")
# 
# colnames(batch1.counts$`Antibody Capture`) <- gsub(x = colnames(batch1.counts$`Antibody Capture`), pattern = "-1", replacement = "")
# 
# batch1.counts$`Antibody Capture` <- batch1.counts$`Antibody Capture`[-3,]
# 
# # Create Seurat object from Gene Expression
# batch1 <- CreateSeuratObject(counts = batch1.counts$`Gene Expression`)
# 
# # Create hashtag Assay object
# hashtag_assay <- CreateAssayObject(counts=batch1.counts$`Antibody Capture`)
# 
# # Add to Seurat object
# batch1[["HTO"]] <- hashtag_assay
# 
# # Rename rows to sample names
# rownames(batch1[["HTO"]]) <- c("1180_CD4_M24", "1344_CD4_S51", "1344_CD4_S174", "1180_CD8_S76", "1180_CD8_ORF3a-28", "1134_CD8_NP1", "1344_CD8_NP13")
# 
# DefaultAssay(batch1) <- "RNA"
# 
# batch1 <- NormalizeData(batch1)
# batch1 <- FindVariableFeatures(batch1)
# batch1 <- ScaleData(batch1)
# batch1 <- RunPCA(batch1, verbose = FALSE)
# batch1 <- FindNeighbors(batch1, dims = 1:30)
# batch1 <- FindClusters(batch1, resolution = 0.8, verbose = FALSE)
# batch1 <- RunUMAP(batch1, dims = 1:30)
# DimPlot(batch1)


```

