---
title: "TCR CDR3 Clustering"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tcR)
library(kernlab)
library(igraph)
library(stringdist)
library(tidyverse)
library(dendextend)
library(factoextra)

```

TCR similarity/dissimilarity by hierarchical clustering of CDR3 sequences

## Read in all data (all beta cells)
```{r, include=FALSE}
# # load in data and create mixcr_b object
# setwd('/t1-data/user/lfelce/TCR_analysis/cd8_np16_new/')
# # create list of file names
# trb_list <- list.files(path = ".", recursive = TRUE,
#                         pattern = "\\TRB.txt$",
#                         full.names = TRUE)
# trb_list <- trb_list %>% str_replace("./*", "")
# # parse in files with tcR
# mixcr_b_np16 <- parse.file.list(trb_list, "mixcr")
# # load in data and create  mixcr_b object
# setwd('/t1-data/user/lfelce/TCR_analysis/cd8_orf_new/')
# # create list of file names, remove minibulk
# trb_list <- list.files(path = ".", recursive = TRUE,
#                        pattern = "\\TRB.txt$",
#                        full.names = TRUE)
# trb_list <- trb_list %>% str_replace("./*", "")
# trb_list <-trb_list[!str_detect(trb_list,pattern="minibulk")]
# # parse in files with tcR
# mixcr_b_orf3a <- parse.file.list(trb_list, "mixcr")
# # combine mixcr files
# mixcr_b <- c(mixcr_b_np16, mixcr_b_orf3a)
# # sort alphabetically
# mixcr_b <- mixcr_b[order(names(mixcr_b))]
# # create list of cell numbers and cell names
# # different lengths so same cell will be different number in a or b
# mixcr_b_names <- as.data.frame(names(mixcr_b))
# mixcr_b_names <- tibble::rownames_to_column(mixcr_b_names, "cell_number")
# # rename columns
# colnames(mixcr_b_names) <- c("cell_number", "cell_name")
# # mixcr_b
# datalist = list()
# for (i in (1:length(mixcr_b))) {
#   dat <- data.frame(c(mixcr_b[[i]][3], mixcr_b[[i]][4], mixcr_b[[i]][6], mixcr_b[[i]][7], mixcr_b[[i]][8]))
#   dat$i <- i # keep track of which iteration produced it
#   datalist[[i]] <- dat # add it to list
# }
# # combine columns for each cell
# # 1 beta cell only
# big_data = do.call(rbind, datalist)
# trb <- big_data %>% group_by(i) %>% filter(n() == 1)
# colnames(trb) <- c("clone_count", "clone_fraction","CDR3_beta", "TRBV", "TRBJ", "cell_number")
# trb <- merge(trb, mixcr_b_names, by="cell_number")
# 
# # tidy up
# trb <- mutate(trb, beta=paste(TRBV, TRBJ, sep="_"))
# 
# trb$epitope <- ifelse(grepl("NP16|B7_SPR", trb$cell_name), "NP16",
#                  ifelse(grepl("ORF3a-28", trb$cell_name), "ORF3a-28", ""))
# 
# 
# # subset epitopes
# np16_sc <- trb[!grepl("ORF3a-28", trb$epitope),]
# np16_sc <- np16_sc[,c("cell_name", "CDR3_beta", "beta")]
# 
# orf3a_sc <- trb[!grepl("NP16", trb$epitope),]
# orf3a_sc <- orf3a_sc[,c("cell_name", "CDR3_beta", "beta")]
# 

# clones
np16_clones <- read.csv("/t1-data/user/lfelce/TCR_analysis/tcr_clones_results/cd8_np16_tcr_clones.csv", header=TRUE)
np16_clones <- np16_clones[!duplicated(np16_clones$CloneName),]
np16_clones <- np16_clones[,c("CloneName", "CDR3_aa.y", "beta")]
colnames(np16_clones) <- c("cell_name", "CDR3_beta", "beta")

orf3a_clones <- read.csv("/t1-data/user/lfelce/TCR_analysis/tcr_clones_results/cd8_orf3a-28_tcr_clones.csv", header=TRUE)
orf3a_clones <- orf3a_clones[!duplicated(orf3a_clones$CloneName),]
orf3a_clones <- orf3a_clones[,c("CloneName", "CDR3_aa.y", "beta")]
colnames(orf3a_clones) <- c("cell_name", "CDR3_beta", "beta")

# all beta cells
all_beta <- read.csv("/t1-data/user/lfelce/TCR_analysis/new_mixcr_results/all_beta.csv")

np16_cells <- all_beta[!grepl("M24|S34|ORF3a-28",all_beta$epitope),]
np16_cells <- mutate(np16_cells, beta=paste(TRBV, TRBJ, sep="_"))
np16_cells <- np16_cells[,c("cell_name", "CDR3_beta", "beta")]

orf3a_cells <- all_beta[!grepl("M24|S34|NP16",all_beta$epitope),]
orf3a_cells <- mutate(orf3a_cells, beta=paste(TRBV, TRBJ, sep="_"))
orf3a_cells <- orf3a_cells[,c("cell_name", "CDR3_beta", "beta")]

# combine clones and single cell data
np16 <- rbind(np16_cells, np16_clones)
orf3a <- rbind(orf3a_cells, orf3a_clones)

# add  metadata
np16$patient <- ifelse(grepl("005", np16$cell_name), "005",
                          ifelse(grepl("1131-TP-1",np16$cell_name), "1131-TP-1",
                          ifelse(grepl("1131-TP-2|1131-CL", np16$cell_name), "1131-TP-2",
                          ifelse(grepl("1153", np16$cell_name), "1153",
                          ifelse(grepl("1201", np16$cell_name), "1201","")))))

np16$severity <- ifelse(grepl("005", np16$cell_name), "mild",
                          ifelse(grepl("1131-TP-1", np16$cell_name), "severe",
                          ifelse(grepl("1131-TP-2|1131-CL", np16$cell_name), "severe",
                          ifelse(grepl("1153", np16$cell_name), "severe",
                          ifelse(grepl("1201", np16$cell_name), "mild","")))))

np16$cell_type <- ifelse(grepl("CL", np16$cell_name), "clone", "single cell")


orf3a$patient <- ifelse(grepl("1105", orf3a$cell_name), "1105",
                          ifelse(grepl("1134-TP1", orf3a$cell_name), "1134-TP-1",
                          ifelse(grepl("1134-TP-2|1134-TP2", orf3a$cell_name), "1134-TP-2",
                          ifelse(grepl("1525-TP-1|1525-TP1", orf3a$cell_name), "1525-TP-1",
                          ifelse(grepl("1525-TP-2|1525-TP2", orf3a$cell_name), "1525-TP-2","")))))

orf3a$severity <- "severe"

orf3a$cell_type <- ifelse(grepl("CL", orf3a$cell_name), "clone", "single cell")

orf3a <- orf3a[-726,]

```


## Distance matrix using stringdist and hierarchical clustering
Hierarchical clustering from distance matrix (dendrogram)
### CD8 NP16 beta

```{r}
# generate distance matrix
d <- stringdistmatrix(np16$CDR3_beta, np16$CDR3_beta, method="lv")
rownames(d) <- np16$cell_name
colnames(d) <- np16$patient

# hierarchical clustering 
cl <- hclust(dist(d, method="euclidean"), method="complete")

dend <- as.dendrogram(cl)

# labels as patient number, colour as cell type
correct_labels <- cl$labels[cl$order]
correct_labels <- as.data.frame(correct_labels)
correct_labels$patient <- ifelse(grepl("005", correct_labels$correct_labels), "005",
                          ifelse(grepl("1131-TP-1",correct_labels$correct_labels), "1131-TP-1",
                          ifelse(grepl("1131-TP-2|1131-CL", correct_labels$correct_labels), "1131-TP-2",
                          ifelse(grepl("1153", correct_labels$correct_labels), "1153",
                          ifelse(grepl("1201", correct_labels$correct_labels), "1201","")))))

correct_labels$cell_type <- ifelse(grepl("CL", correct_labels$correct_labels), "clone", "single cell")

labels(dend) <- correct_labels$patient


library(colorspace)
cell_type <- as.factor(np16$cell_type)

n_cell_type <- length(unique(cell_type))
cols_2 <- colorspace::rainbow_hcl(n_cell_type)
col_cell_type <- cols_2[cell_type]

labels_colors(dend) <- col_cell_type[order.dendrogram(dend)]

# extract info
np16_info <- as.data.frame(cutree(cl,k=10))

# for saving dendrogram info later (csv)
cells <- cbind(correct_labels, np16_info)

```

```{r, fig.width=40, fig.height=10}
# par(mar = c(1,10,1,10))

dend %>% set("labels_cex",0.5) %>% color_branches(k=10) %>%
  plot(main = "CD8+ NP16 CDR3 beta clustering", cex.main=2)
dend_list <- get_subdendrograms(dend, 10)
par(mfrow = c(2,5))
sapply(dend_list, plot)
legend("topright", legend = levels(cell_type), fill = cols_2, cex=2)

```


### CD8 ORF3a beta
```{r}
d <- stringdistmatrix(orf3a$CDR3_beta, orf3a$CDR3_beta, method="lv")
rownames(d) <- orf3a$cell_name
colnames(d) <- orf3a$patient
cl <- hclust(dist(d, method="euclidean"), method="complete")

dend <- as.dendrogram(cl)

correct_labels <- cl$labels[cl$order]
correct_labels <- as.data.frame(correct_labels)
correct_labels$patient <- ifelse(grepl("1105", correct_labels$correct_labels), "1105",
                          ifelse(grepl("1134-TP1",correct_labels$correct_labels), "1134-TP-1",       
                          ifelse(grepl("1134-TP-2|1134-TP2",correct_labels$correct_labels), "1134-TP-2",
                          ifelse(grepl("1525-TP-1|1525-TP1", correct_labels$correct_labels), "1525-TP-1",
                          ifelse(grepl("1525-TP-2|1525-TP2", correct_labels$correct_labels), "1525-TP-2","")))))
correct_labels$cell_type <- ifelse(grepl("CL", correct_labels$correct_labels), "clone", "single cell")

labels(dend) <- correct_labels$patient  

library(colorspace)
cell_type <- as.factor(orf3a$cell_type)

n_cell_type <- length(unique(cell_type))
cols_2 <- colorspace::rainbow_hcl(n_cell_type)
col_cell_type <- cols_2[cell_type]

labels_colors(dend) <- col_cell_type[order.dendrogram(dend)]

# extract info
orf3a_info <- as.data.frame(cutree(cl,k=10))

# for saving later
cells2 <- cbind(correct_labels, orf3a_info)

cells3 <- rbind(cells, cells2)

```

```{r, fig.width=40, fig.height=10}
# par(mar = c(10,1,1,0.1))


dend %>% set("labels_cex",0.5) %>% color_branches(k=10) %>%
  plot(main = "CD8+ ORF3a-28 CDR3 beta clustering", cex.main=2)
dend_list <- get_subdendrograms(dend, 10)

par(mfrow = c(2,5))
sapply(dend_list, plot)
legend("topright", legend = levels(cell_type), fill = cols_2, cex=2)

```

Export cell/clone order to CSV
```{r}
write.csv(cells3, "/t1-data/user/lfelce/TCR_analysis/diversity/cd8_np16_orf3a_all_sc_clones_dendrograms.csv")
```


### Combined
```{r}
# # generate distance matrix
# d <- stringdistmatrix(combined2$CDR3_beta, combined2$CDR3_beta, method="lv")
# rownames(d) <- combined2$cell_name
# colnames(d) <- combined2$patient
# cl <- hclust(as.dist(d))
# 
# dend <- as.dendrogram(cl)
# 
# correct_labels <- cl$labels[cl$order]
# correct_labels <- as.data.frame(correct_labels)
# correct_labels$patient <- ifelse(grepl("005", correct_labels$correct_labels), "005",
#                           ifelse(grepl("1131-TP-1",correct_labels$correct_labels), "1131-TP-1",
#                           ifelse(grepl("1131-TP-2", correct_labels$correct_labels), "1131-TP-2",
#                           ifelse(grepl("1153", correct_labels$correct_labels), "1153",
#                           ifelse(grepl("1201", correct_labels$correct_labels), "1201",
#                           ifelse(grepl("1105", correct_labels$correct_labels), "1105",
#                           ifelse(grepl("1134-TP-2",correct_labels$correct_labels), "1134-TP-2",
#                           ifelse(grepl("1525-TP-1", correct_labels$correct_labels), "1525-TP-1",
#                           ifelse(grepl("1525-TP-2", correct_labels$correct_labels), "1525-TP-2",
#                                  "")))))))))
# labels(dend) <- correct_labels$patient
# 
# library(colorspace)
# epitope <- as.factor(combined2$epitope)
# 
# n_epitope <- length(unique(epitope))
# cols_2 <- colorspace::rainbow_hcl(n_epitope)
# col_epitope <- cols_2[epitope]
# 
# labels_colors(dend) <- col_epitope[order.dendrogram(dend)]
```

```{r, fig.width=40, fig.height=8} 
# # par(mar = c(10,1,1,0.1))
# dend %>% 
#     set("labels_cex",1.5) %>% 
#     color_branches(dend, k=4) %>%
#     plot(main = "CD8+ NP16 and ORF3a-28 CDR3 beta clustering", cex.main=3)
# legend("topright", legend = levels(epitope), fill = cols_2, cex=2)
```

## K-means clustering
Not completely unsupervised as have to find and input optimum number of clusters generated. (also won't get a nice dendrogram)
https://uc-r.github.io/kmeans_clustering
### CD8 NP16 beta
```{r, fig.width=5, fig.height=3}
# # generate distance matrix
# d <- stringdistmatrix(np16$CDR3_beta, np16$CDR3_beta, method="lv")
# rownames(d) <- np16$cell_name
# colnames(d) <- np16$patient
# df <- as.matrix(get_dist(d))
# 
# # find optimum number of clusters
# # elbow method
# set.seed(123)
# fviz_nbclust(df, kmeans, method = "wss")
# 
# # average silhouette method
# set.seed(123)
# fviz_nbclust(df, kmeans, method = "silhouette")

```


```{r, fig.height=10, fig.width=10}
# k2 <- kmeans(df, centers = 10, nstart = 25)
# 
# fviz_cluster(k2, data = df, labelsize=6)

```
### CD8 ORF3a-28 beta

```{r, fig.width=5, fig.height=3}
# # generate distance matrix
# d <- stringdistmatrix(orf3a$CDR3_beta, orf3a$CDR3_beta, method="lv")
# rownames(d) <- orf3a$cell_name
# colnames(d) <- orf3a$patient
# df <- as.matrix(get_dist(d))
# 
# # find optimum number of clusters
# # elbow method
# set.seed(123)
# fviz_nbclust(df, kmeans, method = "wss")
# 
# # average silhouette method
# set.seed(123)
# fviz_nbclust(df, kmeans, method = "silhouette")

```


```{r, fig.height=10, fig.width=10}
# k2 <- kmeans(df, centers = 10, nstart = 25)
# 
# fviz_cluster(k2, data = df, labelsize=6)

```





## CDR3 clustering (Benny's code)
This script creates and plots networks of TCRs which share sequence similarity as measured by the string kernel, looking for similar triplets. Code from Benny Chain (UCL)

### CD8 NP16 beta
```{r, fig.height=6, fig.width=6}
# # this script requires two sets of input; a set of  expanded (>2) CDR3 sequences (each CDR3 should only occur once) in the set of TCRs to be analysed, 
# # in a variable called CDR_exp; and all the CDR3s (each CDR3 should only occur once) from the same set, in a variable called CDR_all.    
# 
# # should be character list not dataframe
# CDR_exp <- np16 %>% group_by(CDR3_beta) %>% filter(n() >= 2)
# CDR_exp <- unique(CDR_exp$CDR3_beta)
# 
# CDR_all <- unique(np16$CDR3_beta)
# 
# #nc<-c()
# #network_sum<-c()
# 
# # initialize the string kernel, which measures the pairwise distance between string vectors, and outputs a kernel matrix 
# # containing the string distance between each CDR3 in set 1 and each CDR3 in set 2.  
# sk <- stringdot(type="spectrum", length=3, normalized=TRUE)
# 
# # calculate string distance between expanded clones and rest
# km_CDR <- kernelMatrix(sk,CDR_exp,CDR_all)
# 
# # remove all the CDR3s which are remotely connected to any of the inputs
# colmax <- apply(km_CDR,2,max)
# # optionally plot a histogram of distances
# # hist(colmax)
# 
# # chose top 1500 TCRS and discard the rest. This is an arbitrary cutoff, which retains all the similar TCRs
# # but does not produce too large a distance matrix 
# l_colmax <- min(1500,length(colmax))
# rem <- order(colmax,decreasing = TRUE)[1:l_colmax]
# 
# # now work out full kernel  matrix for the remaining 1500 CDRs
# km_CDR_all<-kernelMatrix(sk,CDR_all[rem],CDR_all[rem])
# 
# #replace those > 0.98 with 0; this removes any identical pairs of CDR3s. 
# ones <- which(km_CDR_all>0.98)
# km_CDR_all[ones]<-0
# 
# # optionally plot a histogram of distances again to check 
# # hist(km_CDR_all)
# 
# # convert the distance matrix into an adjacentcy  matrix; any two TCRs which are closer than 
# # an arbitrary threshold th, are given a distance value of 1.  Any which are less than th, are given a distance of 0.
# # the size of the threshold is arbitrary, and needs to be set so a suitable control set of CDR3s  give few clusters.
# 
# thc <- 0.5
# connect <- (km_CDR_all >= thc)
# km_CDR_adj <- km_CDR_all
# km_CDR_adj[which(connect)] <- 1
# km_CDR_adj[which(!connect)] <- 0
# 
# # nc is the number of non-zero entries in the set; i.e. number of edges     
# nc <- length(which(km_CDR_adj>0))/2
# 
# # hist(km_CDR_adj)
# # convert the adjacency  matrix into a graph using igraph
# km_graph<-graph.adjacency(km_CDR_adj, mode="undirected",weighted = TRUE,add.colnames=NULL,diag=FALSE)
# summary(km_graph)
# 
# # names of vertices are names of CDR3s
# V(km_graph)$name<-CDR_all[rem]
# 
# # remove isolated nodes; new smaller graph is called km_graph_s
# isolates <- function(g){return(which(igraph::degree(g)==0))}
# km_graph_s<-delete.vertices(km_graph,isolates(km_graph))
# summary(km_graph_s)
# 
# #define which vertices are also exp clones
# inter <- match(CDR_exp,V(km_graph_s)$name)
# exp_v <- inter[which(!is.na(inter))]
# 
# names_g<-V(km_graph_s)$name
# 
# # collect summary data from the graph
# # number of vertices
# n_nodes1 <- length(V(km_graph_s))
# 
# # size of each cluster
# cluster_size <- clusters(km_graph_s)$csize
# 
# # number of clusters
# noclusters <- clusters(km_graph_s)$no
# 
# # degree of each vertix 
# degree <- igraph::degree(km_graph_s,V(km_graph_s))
# degree_dist <- degree.distribution(km_graph_s)
# 
# # which node(vertex)  is in which cluster 
# membership <- clusters(km_graph_s)$membership
# 
# # capture network layout so can plot reproducibly
# coords.exp <- layout.kamada.kawai(km_graph_s)
# 
# # Create the plot
# plot(km_graph_s, vertex.label=NA,layout=coords.exp,edge.width=1,vertex.color="black",edge.color="grey",vertex.size=4)
# title(main="Expanded CDR3s from CD8+ NP16 CDR3 beta", cex.main=1, font.main=1)
# 
# #can add text to plot using mtext
# #mtext(paste0(nam," (thr = ",thc," c.thr = ",c.thr,")"), outer = TRUE, cex = 1)
```


### CD8 ORF3a beta
```{r, fig.height=6, fig.width=6}

# # this script requires two sets of input; a set of  expanded CDR3 sequences (each CDR3 should only occur once) in the set of TCRs to be analysed, 
# # in a variable called CDR_exp; and all the CDR3s (each CDR3 should only occur once) from the same set, in a variable called CDR_all.    
# 
# # should be character list not dataframe
# CDR_exp <- orf3a %>% group_by(CDR3_beta) %>% filter(n() >= 2)
# CDR_exp <- unique(CDR_exp$CDR3_beta)
# 
# CDR_all <- unique(orf3a$CDR3_beta)
# 
# #nc<-c()
# #network_sum<-c()
# 
# # initialize the string kernel, which measures the pairwise distance between string vectors, and outputs a kernel matrix 
# # containing the string distance between each CDR3 in set 1 and each CDR3 in set 2.  
# sk <- stringdot(type="spectrum", length=3, normalized=TRUE)
# 
# # calculate string distance between expanded clones and rest
# km_CDR <- kernelMatrix(sk,CDR_exp,CDR_all)
# 
# # remove all the CDR3s which are remotely connected to any of the inputs
# colmax <- apply(km_CDR,2,max)
# # optionally plot a histogram of distances
# # hist(colmax)
# 
# # chose top 1500 TCRS and discard the rest. This is an arbitrary cutoff, which retains all the similar TCRs
# # but does not produce too large a distance matrix 
# l_colmax <- min(1500,length(colmax))
# rem <- order(colmax,decreasing = TRUE)[1:l_colmax]
# 
# # now work out full kernel  matrix for the remaining 1500 CDRs
# km_CDR_all<-kernelMatrix(sk,CDR_all[rem],CDR_all[rem])
# 
# #replace those > 0.98 with 0; this removes any identical pairs of CDR3s. 
# ones <- which(km_CDR_all>0.98)
# km_CDR_all[ones]<-0
# 
# # optionally plot a histogram of distances again to check 
# # hist(km_CDR_all)
# 
# # convert the distance matrix into an adjacentcy  matrix; any two TCRs which are closer than 
# # an arbitrary threshold th, are given a distance value of 1.  Any which are less than th, are given a distance of 0.
# # the size of the threshold is arbitrary, and needs to be set so a suitable control set of CDR3s  give few clusters.
# 
# thc <- 0.5
# connect <- (km_CDR_all >= thc)
# km_CDR_adj <- km_CDR_all
# km_CDR_adj[which(connect)] <- 1
# km_CDR_adj[which(!connect)] <- 0
# 
# # nc is the number of non-zero entries in the set; i.e. number of edges     
# nc <- length(which(km_CDR_adj>0))/2
# 
# # hist(km_CDR_adj)
# # convert the adjacency  matrix into a graph using igraph
# km_graph<-graph.adjacency(km_CDR_adj, mode="undirected",weighted = TRUE,add.colnames=NULL,diag=FALSE)
# summary(km_graph)
# 
# # names of vertices are names of CDR3s
# V(km_graph)$name<-CDR_all[rem]
# 
# # remove isolated nodes; new smaller graph is called km_graph_s
# isolates <- function(g){return(which(igraph::degree(g)==0))}
# km_graph_s<-delete.vertices(km_graph,isolates(km_graph))
# summary(km_graph_s)
# 
# #define which vertices are also exp clones
# inter <- match(CDR_exp,V(km_graph_s)$name)
# exp_v <- inter[which(!is.na(inter))]
# 
# names_g<-V(km_graph_s)$name
# 
# # collect summary data from the graph
# # number of vertices
# n_nodes1 <- length(V(km_graph_s))
# 
# # size of each cluster
# cluster_size <- clusters(km_graph_s)$csize
# 
# # number of clusters
# noclusters <- clusters(km_graph_s)$no
# 
# # degree of each vertix 
# degree <- igraph::degree(km_graph_s,V(km_graph_s))
# degree_dist <- degree.distribution(km_graph_s)
# 
# # which node(vertex)  is in which cluster 
# membership <- clusters(km_graph_s)$membership
# 
# # capture network layout so can plot reproducibly
# coords.exp <- layout.kamada.kawai(km_graph_s)
# 
# # Create the plot
# plot(km_graph_s, vertex.label=NA,layout=coords.exp,edge.width=1,vertex.color="black",edge.color="grey",vertex.size=4)
# title(main="Expanded CDR3s from CD8+ ORF3a beta", cex.main=1, font.main=1)
# 
# #can add text to plot using mtext
# #mtext(paste0(nam," (thr = ",thc," c.thr = ",c.thr,")"), outer = TRUE, cex = 1)
```


