---
title: "TCR Repertoire Diversity"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tcR)
library(data.table)
library(stringr)
library(plyr)
library(dplyr)
library(cowplot)
library(tidyverse)
library(gridExtra)
library(ggplot2)

```

# NP16 and ORF3a-28 single beta cell using Isar's code - REPEAT
Each repertoire is single patient, keep all clones (even non-expanded). Collapse clones so only have diversity indexes for unique clones. Should have diversity index calculated per repertoire NOT for each TCR

## Diversity functions
```{r}
`diversity_function` <-
  function (x, index = "shannon", MARGIN = 1, base = exp(1))
  {
    x <- drop(as.matrix(x))
    if (!is.numeric(x))
      stop("input data must be numeric")
    if (any(x < 0, na.rm = TRUE))
      stop("input data must be non-negative")
    INDICES <- c("shannon", "simpson", "invsimpson")
    index <- match.arg(index, INDICES)
    if (length(dim(x)) > 1) {
      total <- apply(x, MARGIN, sum)
      x <- sweep(x, MARGIN, total, "/")
    } else {
      x <- x/(total <- sum(x))
    }
    if (index == "shannon")
      x <- -x * log(x, base)
    else
      x <- x*x
    if (length(dim(x)) > 1)
      H <- apply(x, MARGIN, sum, na.rm = TRUE)
    else
      H <- sum(x, na.rm = TRUE)
    if (index == "simpson")
      H <- 1 - H
    else if (index == "invsimpson")
      H <- 1/H
    ## check NA in data
    if (any(NAS <- is.na(total)))
      H[NAS] <- NA
    H
  }
```

## Read in data, diversity calculation
Separately for NP16 and ORF3a-28 then combine for plotting
```{r, include=FALSE}
# load in data and create mixcr_b object
setwd('/t1-data/user/lfelce/TCR_analysis/cd8_np16_new/')

# create list of file names
trb_list <- list.files(path = ".", recursive = TRUE,
                        pattern = "\\TRB.txt$",
                        full.names = TRUE)
trb_list <- trb_list %>% str_replace("./*", "")

# parse in files with tcR
mixcr_b_np16 <- parse.file.list(trb_list, "mixcr")

# load in data and create  mixcr_b object
setwd('/t1-data/user/lfelce/TCR_analysis/cd8_orf_new/')

# create list of file names, remove minibulk
trb_list <- list.files(path = ".", recursive = TRUE,
                       pattern = "\\TRB.txt$",
                       full.names = TRUE)
trb_list <- trb_list %>% str_replace("./*", "")
trb_list <-trb_list[!str_detect(trb_list,pattern="minibulk")]

# parse in files with tcR
mixcr_b_orf3a <- parse.file.list(trb_list, "mixcr")

# combine mixcr files
mixcr_b <- c(mixcr_b_np16, mixcr_b_orf3a)

# sort alphabetically
mixcr_b <- mixcr_b[order(names(mixcr_b))]

# create list of cell numbers and cell names
# different lengths so same cell will be different number in a or b
mixcr_b_names <- as.data.frame(names(mixcr_b))
mixcr_b_names <- tibble::rownames_to_column(mixcr_b_names, "cell_number")

# rename columns
colnames(mixcr_b_names) <- c("cell_number", "cell_name")

```

```{r}
# mixcr_b
datalist = list()
for (i in (1:length(mixcr_b))) {
  dat <- data.frame(c(mixcr_b[[i]][3], mixcr_b[[i]][4], mixcr_b[[i]][6], mixcr_b[[i]][7], mixcr_b[[i]][8]))
  dat$i <- i # keep track of which iteration produced it
  datalist[[i]] <- dat # add it to list
}
# combine columns for each cell
# 1 beta cell only
big_data = do.call(rbind, datalist)
trb <- big_data %>% group_by(i) %>% filter(n() == 1)
colnames(trb) <- c("clone_count", "clone_fraction","CDR3_beta", "TRBV", "TRBJ", "cell_number")
trb <- merge(trb, mixcr_b_names, by="cell_number")

# tidy up
trb <- mutate(trb, beta=paste(TRBV, TRBJ, sep="_"))

# add metadata
trb$patient <- ifelse(grepl("005", trb$cell_name), "005",
                          ifelse(grepl("1131-TP-1",trb$cell_name), "1131-TP-1",
                          ifelse(grepl("1131-TP-2", trb$cell_name), "1131-TP-2",
                          ifelse(grepl("1153", trb$cell_name), "1153",
                          ifelse(grepl("1201", trb$cell_name), "1201",
                          ifelse(grepl("1105", trb$cell_name), "1105",
                          ifelse(grepl("1134-TP-2",trb$cell_name), "1134",
                          ifelse(grepl("1525-TP-1", trb$cell_name), "1525-TP-1",
                          ifelse(grepl("1525-TP-2", trb$cell_name), "1525-TP-2",
                                 "")))))))))
trb$severity <- ifelse(grepl("005", trb$cell_name), "mild",
                          ifelse(grepl("1131-TP-1",trb$cell_name), "severe",
                          ifelse(grepl("1131-TP-2", trb$cell_name), "severe",
                          ifelse(grepl("1153", trb$cell_name), "severe",
                          ifelse(grepl("1201", trb$cell_name), "mild",
                          ifelse(grepl("1105", trb$cell_name), "severe",
                          ifelse(grepl("1134",trb$cell_name), "severe",
                          ifelse(grepl("1525-TP-1", trb$cell_name), "severe",
                          ifelse(grepl("1525-TP-2", trb$cell_name), "severe",
                                 "")))))))))
trb$timepoint <- ifelse(grepl("TP-1", trb$patient), "acute",
                   ifelse(grepl("005|TP-2|1134|1153|1201|1105", trb$cell_name), "convalescent", ""))

trb$epitope <- ifelse(grepl("NP16|B7_SPR", trb$cell_name), "NP16",
                 ifelse(grepl("ORF3a-28", trb$cell_name), "ORF3a-28", ""))

# subset data by patient 
p005 <- trb[(trb$patient=="005"),]
p1131TP1 <- trb[(trb$patient=="1131-TP-1"),]
p1131TP2 <- trb[(trb$patient=="1131-TP-2"),]
p1153 <- trb[(trb$patient=="1153"),]
p1201 <- trb[(trb$patient=="1201"),]
p1105 <- trb[(trb$patient=="1105"),]
p1134 <- trb[(trb$patient=="1134"),]
p1525TP1 <- trb[(trb$patient=="1525-TP-1"),]
p1525TP2 <- trb[(trb$patient=="1525-TP-2"),]

# create loop to calculate diversity index
patient_list <- list(p005, p1131TP1, p1131TP2, p1153, p1201, p1105, p1134, p1525TP1, p1525TP2)

datalist = list()

for (i in (1:length(patient_list))) {
  table <- table(patient_list[[i]][4])
  div <- diversity_function(table, "shannon")
  datalist[[i]] <- div 
}

combined <-  do.call(rbind, datalist)

combined <- as.data.frame(combined)

colnames(combined) <- "diversity"

combined$patient <- c("005", "1131-TP1", "1131-TP2", "1153", "1201", "1105", "1134", "1525-TP1", "1525-TP2")

combined$severity <- c("mild", "severe", "severe", "severe", "mild", "severe", "severe", "severe", "severe")

combined$epitope <- c("NP16", "NP16", "NP16", "NP16", "NP16", "ORF3a-28", "ORF3a-28", "ORF3a-28", "ORF3a-28")

print(combined)

combined2 <- combined[!grepl("1131-TP1", combined$patient),]

np16 <- combined2[!grepl("ORF3a-28", combined2$epitope),]

```
## Boxplots
```{r}
p1 <- ggplot(combined2, aes(x=epitope, y=diversity, colour=epitope)) +
  geom_boxplot(width=0.5) + geom_jitter(width=0.25) +
  labs(title="Diversity of CD8+ NP16 & ORF3a-28 CDR3 beta",x="Patient", y = "Shannon diversity index") + theme_light()

p2 <- ggplot(np16, aes(x=severity, y=diversity, colour=severity)) +
  geom_jitter(width=0.25) +
  labs(title="Diversity of CD8+ NP16  CDR3 beta",x="Severity", y = "Shannon diversity index") + theme_light()

```

```{r, fig.width=12, fig.height=6}
grid.arrange(p1, p2, ncol=2)
```

## Stats
```{r}
# Welch 2 sample t-test: (y~x) where y is numeric and x is a binary factor
t.test(combined2$diversity~combined2$epitope)

t.test(np16$diversity~np16$severity)

# independent 2-group Mann-Whitney U Test (non-parametric)
# where y is numeric and A is A binary factor
wilcox.test(combined2$diversity~combined2$epitope)

wilcox.test(np16$diversity~np16$severity)

```







