---
title: "TCR CDR3 Clustering"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tcR)
library(kernlab)
library(igraph)
library(stringdist)
library(tidyverse)
library(pheatmap)
library(dendextend)
```


## Read in all data (1 beta cells)
```{r, include=FALSE}
# load in data and create mixcr_b object
setwd('/t1-data/user/lfelce/TCR_analysis/cd8_np16_new/')
# create list of file names
trb_list <- list.files(path = ".", recursive = TRUE,
                        pattern = "\\TRB.txt$",
                        full.names = TRUE)
trb_list <- trb_list %>% str_replace("./*", "")
# parse in files with tcR
mixcr_b_np16 <- parse.file.list(trb_list, "mixcr")
# load in data and create  mixcr_b object
setwd('/t1-data/user/lfelce/TCR_analysis/cd8_orf_new/')
# create list of file names, remove minibulk
trb_list <- list.files(path = ".", recursive = TRUE,
                       pattern = "\\TRB.txt$",
                       full.names = TRUE)
trb_list <- trb_list %>% str_replace("./*", "")
trb_list <-trb_list[!str_detect(trb_list,pattern="minibulk")]
# parse in files with tcR
mixcr_b_orf3a <- parse.file.list(trb_list, "mixcr")
# combine mixcr files
mixcr_b <- c(mixcr_b_np16, mixcr_b_orf3a)
# sort alphabetically
mixcr_b <- mixcr_b[order(names(mixcr_b))]
# create list of cell numbers and cell names
# different lengths so same cell will be different number in a or b
mixcr_b_names <- as.data.frame(names(mixcr_b))
mixcr_b_names <- tibble::rownames_to_column(mixcr_b_names, "cell_number")
# rename columns
colnames(mixcr_b_names) <- c("cell_number", "cell_name")
# mixcr_b
datalist = list()
for (i in (1:length(mixcr_b))) {
  dat <- data.frame(c(mixcr_b[[i]][3], mixcr_b[[i]][4], mixcr_b[[i]][6], mixcr_b[[i]][7], mixcr_b[[i]][8]))
  dat$i <- i # keep track of which iteration produced it
  datalist[[i]] <- dat # add it to list
}
# combine columns for each cell
# 1 beta cell only
big_data = do.call(rbind, datalist)
trb <- big_data %>% group_by(i) %>% filter(n() == 1)
colnames(trb) <- c("clone_count", "clone_fraction","CDR3_beta", "TRBV", "TRBJ", "cell_number")
trb <- merge(trb, mixcr_b_names, by="cell_number")
# tidy up
trb <- mutate(trb, beta=paste(TRBV, TRBJ, sep="_"))
table <- as.data.frame(table(trb$CDR3_beta))
combined2 <- merge(trb, table, by.x="CDR3_beta", by.y="Var1", all.x=TRUE)
combined2$patient <- ifelse(grepl("005", combined2$cell_name), "005",
                          ifelse(grepl("1131-TP-1",combined2$cell_name), "1131-TP-1",
                          ifelse(grepl("1131-TP-2", combined2$cell_name), "1131-TP-2",
                          ifelse(grepl("1153", combined2$cell_name), "1153",
                          ifelse(grepl("1201", combined2$cell_name), "1201",
                          ifelse(grepl("1105", combined2$cell_name), "1105",
                          ifelse(grepl("1134-TP-2",combined2$cell_name), "1134-TP-2",
                          ifelse(grepl("1525-TP-1", combined2$cell_name), "1525-TP-1",
                          ifelse(grepl("1525-TP-2", combined2$cell_name), "1525-TP-2",
                                 "")))))))))
combined2$severity <- ifelse(grepl("005", combined2$cell_name), "mild",
                          ifelse(grepl("1131-TP-1",combined2$cell_name), "severe",
                          ifelse(grepl("1131-TP-2", combined2$cell_name), "severe",
                          ifelse(grepl("1153", combined2$cell_name), "severe",
                          ifelse(grepl("1201", combined2$cell_name), "mild",
                          ifelse(grepl("1105", combined2$cell_name), "severe",
                          ifelse(grepl("1134-TP-2",combined2$cell_name), "severe",
                          ifelse(grepl("1525-TP-1", combined2$cell_name), "severe",
                          ifelse(grepl("1525-TP-2", combined2$cell_name), "severe",
                                 "")))))))))
combined2$timepoint <- ifelse(grepl("TP-1", combined2$patient), "acute",
                   ifelse(grepl("005|TP-2|1153|1201|1105", combined2$cell_name), "convalescent", ""))
combined2$epitope <- ifelse(grepl("NP16|B7_SPR", combined2$cell_name), "NP16",
                 ifelse(grepl("ORF3a-28", combined2$cell_name), "ORF3a-28", ""))


# subset epitopes
np16 <- combined2[!grepl("ORF3a-28", combined2$epitope),]

orf3a <- combined2[!grepl("NP16", combined2$epitope),]


```

## CDR3 clustering (Benny's code)
This script creates and plots networks of TCRs which share sequence similarity as measured by the string kernel, looking for similar triplets. Code from Benny Chain (UCL)

### CD8 NP16 beta
```{r, fig.height=6, fig.width=6}
# this script requires two sets of input; a set of  expanded (>2) CDR3 sequences (each CDR3 should only occur once) in the set of TCRs to be analysed, 
# in a variable called CDR_exp; and all the CDR3s (each CDR3 should only occur once) from the same set, in a variable called CDR_all.    

# should be character list not dataframe
CDR_exp <- np16 %>% group_by(CDR3_beta) %>% filter(n() >= 2)
CDR_exp <- unique(CDR_exp$CDR3_beta)

CDR_all <- unique(np16$CDR3_beta)

#nc<-c()
#network_sum<-c()

# initialize the string kernel, which measures the pairwise distance between string vectors, and outputs a kernel matrix 
# containing the string distance between each CDR3 in set 1 and each CDR3 in set 2.  
sk <- stringdot(type="spectrum", length=3, normalized=TRUE)

# calculate string distance between expanded clones and rest
km_CDR <- kernelMatrix(sk,CDR_exp,CDR_all)

# remove all the CDR3s which are remotely connected to any of the inputs
colmax <- apply(km_CDR,2,max)
# optionally plot a histogram of distances
# hist(colmax)

# chose top 1500 TCRS and discard the rest. This is an arbitrary cutoff, which retains all the similar TCRs
# but does not produce too large a distance matrix 
l_colmax <- min(1500,length(colmax))
rem <- order(colmax,decreasing = TRUE)[1:l_colmax]

# now work out full kernel  matrix for the remaining 1500 CDRs
km_CDR_all<-kernelMatrix(sk,CDR_all[rem],CDR_all[rem])

#replace those > 0.98 with 0; this removes any identical pairs of CDR3s. 
ones <- which(km_CDR_all>0.98)
km_CDR_all[ones]<-0

# optionally plot a histogram of distances again to check 
# hist(km_CDR_all)

# convert the distance matrix into an adjacentcy  matrix; any two TCRs which are closer than 
# an arbitrary threshold th, are given a distance value of 1.  Any which are less than th, are given a distance of 0.
# the size of the threshold is arbitrary, and needs to be set so a suitable control set of CDR3s  give few clusters.

thc <- 0.5
connect <- (km_CDR_all >= thc)
km_CDR_adj <- km_CDR_all
km_CDR_adj[which(connect)] <- 1
km_CDR_adj[which(!connect)] <- 0

# nc is the number of non-zero entries in the set; i.e. number of edges     
nc <- length(which(km_CDR_adj>0))/2

# hist(km_CDR_adj)
# convert the adjacency  matrix into a graph using igraph
km_graph<-graph.adjacency(km_CDR_adj, mode="undirected",weighted = TRUE,add.colnames=NULL,diag=FALSE)
summary(km_graph)

# names of vertices are names of CDR3s
V(km_graph)$name<-CDR_all[rem]

# remove isolated nodes; new smaller graph is called km_graph_s
isolates <- function(g){return(which(igraph::degree(g)==0))}
km_graph_s<-delete.vertices(km_graph,isolates(km_graph))
summary(km_graph_s)

#define which vertices are also exp clones
inter <- match(CDR_exp,V(km_graph_s)$name)
exp_v <- inter[which(!is.na(inter))]

names_g<-V(km_graph_s)$name

# collect summary data from the graph
# number of vertices
n_nodes1 <- length(V(km_graph_s))

# size of each cluster
cluster_size <- clusters(km_graph_s)$csize

# number of clusters
noclusters <- clusters(km_graph_s)$no

# degree of each vertix 
degree <- igraph::degree(km_graph_s,V(km_graph_s))
degree_dist <- degree.distribution(km_graph_s)

# which node(vertex)  is in which cluster 
membership <- clusters(km_graph_s)$membership

# capture network layout so can plot reproducibly
coords.exp <- layout.kamada.kawai(km_graph_s)

# Create the plot
plot(km_graph_s, vertex.label=NA,layout=coords.exp,edge.width=1,vertex.color="black",edge.color="grey",vertex.size=4)
title(main="Expanded CDR3s from CD8+ NP16 CDR3 beta", cex.main=1, font.main=1)

#can add text to plot using mtext
#mtext(paste0(nam," (thr = ",thc," c.thr = ",c.thr,")"), outer = TRUE, cex = 1)
```


### CD8 ORF3a beta
```{r, fig.height=6, fig.width=6}

# this script requires two sets of input; a set of  expanded CDR3 sequences (each CDR3 should only occur once) in the set of TCRs to be analysed, 
# in a variable called CDR_exp; and all the CDR3s (each CDR3 should only occur once) from the same set, in a variable called CDR_all.    

# should be character list not dataframe
CDR_exp <- orf3a %>% group_by(CDR3_beta) %>% filter(n() >= 2)
CDR_exp <- unique(CDR_exp$CDR3_beta)

CDR_all <- unique(orf3a$CDR3_beta)

#nc<-c()
#network_sum<-c()

# initialize the string kernel, which measures the pairwise distance between string vectors, and outputs a kernel matrix 
# containing the string distance between each CDR3 in set 1 and each CDR3 in set 2.  
sk <- stringdot(type="spectrum", length=3, normalized=TRUE)

# calculate string distance between expanded clones and rest
km_CDR <- kernelMatrix(sk,CDR_exp,CDR_all)

# remove all the CDR3s which are remotely connected to any of the inputs
colmax <- apply(km_CDR,2,max)
# optionally plot a histogram of distances
# hist(colmax)

# chose top 1500 TCRS and discard the rest. This is an arbitrary cutoff, which retains all the similar TCRs
# but does not produce too large a distance matrix 
l_colmax <- min(1500,length(colmax))
rem <- order(colmax,decreasing = TRUE)[1:l_colmax]

# now work out full kernel  matrix for the remaining 1500 CDRs
km_CDR_all<-kernelMatrix(sk,CDR_all[rem],CDR_all[rem])

#replace those > 0.98 with 0; this removes any identical pairs of CDR3s. 
ones <- which(km_CDR_all>0.98)
km_CDR_all[ones]<-0

# optionally plot a histogram of distances again to check 
# hist(km_CDR_all)

# convert the distance matrix into an adjacentcy  matrix; any two TCRs which are closer than 
# an arbitrary threshold th, are given a distance value of 1.  Any which are less than th, are given a distance of 0.
# the size of the threshold is arbitrary, and needs to be set so a suitable control set of CDR3s  give few clusters.

thc <- 0.5
connect <- (km_CDR_all >= thc)
km_CDR_adj <- km_CDR_all
km_CDR_adj[which(connect)] <- 1
km_CDR_adj[which(!connect)] <- 0

# nc is the number of non-zero entries in the set; i.e. number of edges     
nc <- length(which(km_CDR_adj>0))/2

# hist(km_CDR_adj)
# convert the adjacency  matrix into a graph using igraph
km_graph<-graph.adjacency(km_CDR_adj, mode="undirected",weighted = TRUE,add.colnames=NULL,diag=FALSE)
summary(km_graph)

# names of vertices are names of CDR3s
V(km_graph)$name<-CDR_all[rem]

# remove isolated nodes; new smaller graph is called km_graph_s
isolates <- function(g){return(which(igraph::degree(g)==0))}
km_graph_s<-delete.vertices(km_graph,isolates(km_graph))
summary(km_graph_s)

#define which vertices are also exp clones
inter <- match(CDR_exp,V(km_graph_s)$name)
exp_v <- inter[which(!is.na(inter))]

names_g<-V(km_graph_s)$name

# collect summary data from the graph
# number of vertices
n_nodes1 <- length(V(km_graph_s))

# size of each cluster
cluster_size <- clusters(km_graph_s)$csize

# number of clusters
noclusters <- clusters(km_graph_s)$no

# degree of each vertix 
degree <- igraph::degree(km_graph_s,V(km_graph_s))
degree_dist <- degree.distribution(km_graph_s)

# which node(vertex)  is in which cluster 
membership <- clusters(km_graph_s)$membership

# capture network layout so can plot reproducibly
coords.exp <- layout.kamada.kawai(km_graph_s)

# Create the plot
plot(km_graph_s, vertex.label=NA,layout=coords.exp,edge.width=1,vertex.color="black",edge.color="grey",vertex.size=4)
title(main="Expanded CDR3s from CD8+ ORF3a beta", cex.main=1, font.main=1)

#can add text to plot using mtext
#mtext(paste0(nam," (thr = ",thc," c.thr = ",c.thr,")"), outer = TRUE, cex = 1)
```

## Pairwise similarity matrix using stringdist

### CD8 NP16 beta
```{r}
datalist = list()
for (i in (1:length(np16$cell_name))) {
  dat <- stringsim(np16$CDR3_beta[i], np16$CDR3_beta, method="lv") 
  datalist[[i]] <- dat # add it to list
}
big_data = do.call(rbind, datalist)

```

### Hierarchal clustering from pairwise similarity matrix (dendrogram)
```{r}
# take pairwise similarity matrix
df <- big_data
rownames(df) <- np16$cell_name
colnames(df) <- np16$patient

# scale, calculate distance (Euclidean) and perform hierarchal clustering
df <- scale(df)
d <- dist(df)
hc <- hclust(d,method="complete")

dend <- as.dendrogram(hc)

patient <- as.factor(np16$patient)
n_patient <- length(unique(patient))
cols_5 <- colorspace::rainbow_hcl(n_patient, c = 70, l  = 50)
col_patient <- cols_5[patient]

labels_colors(dend) <- col_patient[order.dendrogram(dend)]
labels(dend) <- colnames(df)
dend <- set(dend, "labels_cex", 2)

```

```{r, fig.width=40, fig.height=15}
plot(dend, main = "CD8+ NP16 CDR3 beta clustering")
legend("topleft", legend = levels(patient), fill = cols_5)

```

### CD8 ORF3a beta
```{r}
datalist = list()
for (i in (1:length(orf3a$cell_name))) {
  dat <- stringsim(orf3a$CDR3_beta[i], orf3a$CDR3_beta, method="lv") 
  datalist[[i]] <- dat # add it to list
}
big_data = do.call(rbind, datalist)

```

### Hierarchal clustering from pairwise similarity matrix (dendrogram)
```{r}
# take pairwise similarity matrix
df <- big_data
rownames(df) <- orf3a$cell_name
colnames(df) <- orf3a$patient

# scale, calculate distance (Euclidean) and perform hierarchal clustering
df <- scale(df)
d <- dist(df)
hc <- hclust(d,method="complete")

dend <- as.dendrogram(hc)

patient <- as.factor(orf3a$patient)
n_patient <- length(unique(patient))
cols_4 <- colorspace::rainbow_hcl(n_patient, c = 70, l  = 50)
col_patient <- cols_4[patient]

labels_colors(dend) <- col_patient[order.dendrogram(dend)]
labels(dend) <- colnames(df)
dend <- set(dend, "labels_cex", 2)

```

```{r, fig.width=40, fig.height=15}
plot(dend, main = "CD8+ ORF3a CDR3 beta clustering")
legend("topleft", legend = levels(patient), fill = cols_4)

```
